---
title: "Basics of cubic spline models"
author: "Steve Simon"
date: "2021-11-01"
categories:
- Recommendation
tags:
- Linear regression
output: html_document
---

```{r setup}
suppressMessages(suppressWarnings(library(broom)))
suppressMessages(suppressWarnings(library(dplyr)))
suppressMessages(suppressWarnings(library(magrittr)))
```

### Artificial data

Suppose you have some data where you suspect the behavior differs for x=1 to 5, x=6 to 10, x=11 to 15, and x=16 to 20.

```{r}
library(ggplot2)
x <- 1:20
y <- rep(40, 20)
y[6:20] <- y[6:20]-3*(1:15)
y[11:20] <- y[11:20]+7*(1:10)
y[16:20] <- y[16:20]-(1:5)^2
y <- y+round(rnorm(10), 1)
simulated_example <- data.frame(x, y)
simulated_example
ggplot(simulated_example, aes(x, y)) +
  geom_point()
```

A single cubic function for all the data doesn't quite capture what is happening here.

```{r}
extra_values <- data.frame(x=seq(0, 21, length=2200))
simulated_example %<>% full_join(extra_values, by="x")
```

```{r}
single_cubic_model <- lm(y~poly(x, 3), data=simulated_example)
single_cubic_prediction <- augment(single_cubic_model)
ggplot(single_cubic_prediction, aes(x, y)) + 
  geom_point() +
  geom_line(aes(y=.fitted))
```

So, we could fit a cubic model for the first five data points, for the second five, the third five, and the fourth five. This is a bit much: a cubic model has four parameters, so fitting four of them would use up 16 degrees of freedom in a data set with only 20 observations. But bear with me a bit on this.

```{r}
sb1 <- (simulated_example$x<=5)
sb2 <- (simulated_example$x>5)&(simulated_example$x<=10)
sb3 <- (simulated_example$x>10)&(simulated_example$x<=15)
sb4 <- (simulated_example$x>15)

simulated_example %>%
  filter(sb1) %>%
  lm(y~poly(x, 3), data=.) %>%
  augment(newdata=simulated_example[sb1, ]) -> first_cubic_prediction
lm(y~poly(x, 3), subset=sb2, data=simulated_example) %>% 
  augment(newdata=simulated_example[sb2, ]) %>%
  select(x, .fitted) -> second_cubic_prediction
lm(y~poly(x, 3), subset=sb3, data=simulated_example) %>%
  augment(newdata=simulated_example[sb3, ]) %>% 
  select(x, .fitted) -> third_cubic_prediction
lm(y~poly(x, 3), subset=sb4, data=simulated_example) %>%
  augment(newdata=simulated_example[sb4, ]) %>%
  select(x, .fitted) -> fourth_cubic_prediction

first_cubic_prediction %>%
  bind_rows(second_cubic_prediction) %>%
  bind_rows(third_cubic_prediction) %>%
  bind_rows(fourth_cubic_prediction) %>%
  full_join(simulated_example, by="x") -> combined_prediction
ggplot(combined_prediction, aes(x, y.y)) +
  geom_point() +
  geom_line(aes(y=.fitted))
```

This function is not continuous or smooth. To make a single smooth continuous function you need to define

```{r}
simulated_example %>%
  mutate(x1=x) %>%
  mutate(x2=x^2) %>%
  mutate(x3=x^3) %>%
  mutate(x4=(x>5)*(x-5)^3) %>%
  mutate(x5=(x>10)*(x-10)^3) %>%
  mutate(x6=(x>15)*(x-15)^3) -> spline_basis
```

```{r}
lm(y~x1+x2+x3+x4+x5+x6, data=spline_basis) %>%
  augment(newdata=spline_basis) %>%
  ggplot(aes(x, y)) +
    geom_point() +
  geom_line(aes(y=.fitted))
```

