---
title: "A systematic review shows no performance benefit of machine learning over logistic regression for clinical prediction models"
author: "Steve Simon"
source: "http://blog.pmean.com/machine-learning-vs-logistic-regression/"
date: "2019-03-15"
categories:
- Recommendation
tags:
- Data science
- Logistic regression
- Systematic overviews
output: html_document
---

### Downward bend

Suppose you want to fit a function that is flat from x=a to x=b and which declines linearly from x=b to x=c.

Here is the formula:

$f(x) = d - e*(x-b)_+$

where 

$u_+ = u$ if $u \ge 0$, $= 0$ if $u<0$

Here is some simulated data for testing this function.

```{r, echo=FALSE}
library(ggplot2)
x <- 1:20
y <- c(rep(8, 15), 7:3)+round(rnorm(10), 1)
d <- data.frame(x, y)
d
ggplot(d, aes(x, y)) +
  geom_point()
```

```{r}
par(mar=rep(0.1,4))
plot(c(-1,2), 0:1, type="n", axes=FALSE, xlab="", ylab="")
segments(-1, 0, 2, 0)
segments(-1, 1, 2, 1)
segments(-1, 0.6, 0, 0.6)
segments(-1, 0.9, 0, 0.9)
segments( 1, 0.4, 2, 0.4)
segments( 1, 0.8, 2, 0.8)
segments( 0, 0.35, 1, 0.35)
segments( 0, 0.95, 1, 0.95)
x <-seq(0,1, length=100)
y <- 3*x^2-2*x^3
lines(x, 0.35+0.05*y)
lines(x, 0.5+0.05*y)
lines(x, 0.5+0.3*y)
lines(x, 0.6+0.3*y)
lines(x, 0.6-0.25*y)
lines(x, 0.65-0.25*y)
lines(x, 0.65-0.1*y)
lines(x, 0.85-0.15*y)
lines(x, 0.85+0.05*y)
lines(x, 0.9-0.15*y)
lines(x, 0.9+0.05*y)
lines(x, 0.9-0.15*y)
lines(x, 0.95-0.15*y)
```

```{r}
x <- seq(0,1, length=100)
y <- 3*x^2-2*x^3
plot(x, y)
```

Aris Perperoglou, Willi Sauerbrei, Michal Abrahamowicz, Matthias Schmid. A review of spline function procedures in R. BMC Medical Research Methodology, 2019-03-06, 19(46). Available in [html format][per1] or [pdf format][per2].


[per1]: https://bmcmedresmethodol.biomedcentral.com/articles/10.1186/s12874-019-0666-3
[per2]: https://bmcmedresmethodol.biomedcentral.com/track/pdf/10.1186/s12874-019-0666-3.pdf
