---
title: "The hedging hyperprior for the normal distribution: known mean, unknown precision"
source: "New"
date: "2023-03-14"
categories:
- Blog post
tags:
- Bayesian statistics
- Rmarkdown graphs
output: html_document
page_update: complete
---

```{r normal-hedging-prior-20}
suppressMessages(suppressWarnings(library(MASS)))
suppressMessages(suppressWarnings(library(tidyverse)))
knitr::opts_chunk$set(echo=TRUE, fig.width=8, fig.height=2)
```

I want to write a paper about the hedging hyperprior. I need to work out some simple examples first. Here is an example involving the normal distribution.

<!---more--->

### A simple Bayesian analysis with an informative prior.

There are lots of guides on the Internet on using a normal prior for data that is normally distributed with the population standard deviation $\sigma$ known or unknown. I used papers written by [Kevin P. Murphy][mur1] and [Michael I. Jordan][jor1] to get these formulas.

For simplicity, all sums and products are assumed to go from i=1 to n.

Consider a Bayesian analysis of an outcome variable that has a normal distribution with known mean and unknown precision. Assume that both the mean and the precision of the normal distribution are unknown. Other cases with [mean unknown and precision known][sim3] and with [both mean and precision unknown][sim5] appear on a different web pages.

[sim3]: http://new.pmean.com/normal-hedging-hyperprior-part1
[sim5]: http://new.pmean.com/normal-hedging-hyperprior-part3

Assume that the mean of the normal distribution, $\mu$, is known and equal to $\mu_F$. An earlier case with precision known and the mean  unknown appears on a different web page. A more general case, with both the mean and the precision of the normal distribution unknown is also described on a separate web page. For simplicity, we will drop any multiplicative constants not involving X or $\tau$

$f(X; \tau) \propto
  \sqrt{\tau}
  e^{-\frac{1}{2}\tau(X-\mu_F)^2}$

The conjugate prior for a normal distribution is a gamma distribution.

$g(\tau; \alpha_0, \beta_0) =
  \frac
    {\beta_0^{\alpha_0}}
    {\Gamma(\alpha_0)}
  \tau^{\alpha_0-1}
  e^{-\beta_0 \tau}$

Set the prior parameters to

$\alpha_0=n_0$ 

$\beta_0=n_0\tau_0$

and remove any multiplicative constants with respect to $\tau$ to get

$g(\tau; n_0, \tau_0) \propto \tau^{n_0-1} e^{-n_0 \tau_0 \tau}$

In this parameterization, $n_0$ is the prior sample size. 

You have collected a sample of n data points, $\mathbf{X} = X_1, X_2, ..., X_n$. This gives you a likelihood of 

$L(\mathbf{X} | \mu) \propto
  \tau^{n/2}
  e^{- \frac{1}{2} \tau\sum(X_i-\mu_0)^2}$

The posterior distribution is proportional to the product of the prior and the likelihood.

$h(\tau | \mathbf{X}) \propto
  \tau^{n_0-1}
  e^{-n_0 \tau / \tau0}
  \tau^{n/2}
  e^{- \frac{1}{2} \tau\sum(X_i-\mu_0)^2}$

which can be rewritten as

$h(\tau | \mathbf{X}) \propto
  \tau^{n/2 + n_0-1}
  e^{-(n_0 \tau0 + \frac{1}{2}\sum(X_i-\mu_0)^2)}$

which is a gamma distribution with posterior parameters

$\alpha_1 = n/2+n_0$ and

$\beta_1 = n_0 \tau_0 + \frac{1}{2}\sum(X_i-\mu_0)^2)$

### A simple numerical example

```{r normal-hedging-prior-21}
x_max <- 8

n0 <- 4
tau0 <- 0.5

alpha0 <- n0
beta0 <- n0*tau0

muF <- 2
n <- 10

x <- muF+ round(rnorm(n), 1)/tau0

ss <- sum((x-muF)^2)

alpha1 <- alpha0 + n
beta1 <- beta0 + 0.5*ss

x_grid <- seq(0.01, x_max, length=1000)

f <- function(x, mu, tau) {
  sqrt(tau/2*pi)*exp(-0.5*tau*(x-mu)^2)
}

g <- function(x, alpha, beta) {
  beta^alpha/gamma(alpha)*x^(alpha-1)*exp(-beta*x)
}

prior05 <- qgamma(0.025, alpha0, beta0) 
prior95 <- qgamma(0.975, alpha0, beta0) 

post05 <- qgamma(0.025, alpha1, beta1)  
post95 <- qgamma(0.975, alpha1, beta1) 

```

Let's illustrate this with an informative prior with $n_0$ = `r n0` and $\tau_0$ = `r tau0`.

```{r normal-hedging-prior-22}
g_grid <- g(x_grid, alpha0, beta0)
outside <- x_grid < prior05 | x_grid > prior95
x_color <- 
  ifelse(outside, "white", "gray")
data.frame(x=x_grid, y=g_grid) %>%
  ggplot(aes(x, y)) +
    geom_segment(aes(xend=x, yend=0), color=x_color) +
    geom_line() + xlab(" ") + ylab(" ")
```

This corresponds to a fairly informative prior distribution. This distribution places $\tau$ within `r sprintf("%.2f", prior05)` and `r sprintf("%.2f", prior95)` with 95% probability.

Let's see what happens with this informative prior if you observe `r n` values of x: `r paste0(x, collapse=", ")`. The likelihood is

```{r normal-hedging-prior-23}
f_grid <- rep(NA, length(x_grid))
for (i in 1:length(x_grid)) {
  f_grid[i] <- prod(f(x, muF, x_grid[i]))
}
x_color <- "white"
data.frame(x=x_grid, y=f_grid) %>%
  ggplot(aes(x, y)) + 
    geom_line() + xlab(" ") + ylab(" ")
```

Multiply these two together to get the posterior distribution.

```{r normal-hedging-prior-24}
h_grid <- g(x_grid, alpha1, beta1)
outside <- x_grid < post05 | x_grid > post95
x_color <- 
  ifelse(outside, "white", "gray")
data.frame(x=x_grid, y=h_grid) %>%
  ggplot(aes(x, y)) +
    geom_segment(aes(xend=x_grid, yend=0), color=x_color) +
    geom_line() + xlab(" ") + ylab(" ")
```

The posterior distribution is not close to either the prior distribution or the likelihood. It places $\tau$ within `r sprintf("%.2f", post05)` and `r sprintf("%.2f", post95)` with 95% probability. It excludes the prior mean of `r tau0` and it excludes `r ss/(n-1)`, the mean of the data.

### The hedging hyperprior

When the likelihood of your data and your informative prior are in conflict, you may wish to downweight the informative prior. You can do this by including a hedging hyperprior.

Define your prior distribution as a gamma distributions with prior 

mean $\tau_0$ and prior precision

$\alpha = 1 + \lambda(n0-1)$ 

$\beta = \alpha \tau_0 $

where $\lambda$ ranges between 0 and 1. When $\lambda=0$, the prior precision reduces to $alpha=1$ which is equivalent to a very diffuse prior. When $lambda=1$, the prior precision is $\alpha=\alpha_0$ or the full strength prior.

Place a hyperprior on the parameter $\lambda$. The uniform(0,1) distribution is a reasonable choice for the hyperprior, but others are possible.

```{r normal-hedging-prior-25}
knitr::opts_chunk$set(echo=TRUE, fig.width=8, fig.height=8)
```

```{r normal-hedging-prior-26}
lambda_grid <- seq(0, 1, length=60)
tau_grid <- seq(0, x_max, length=60)


gstar <- function(tau, lambda) {g(tau, 1+lambda*(alpha0-1), (1+lambda*(alpha0-1))*beta0)}
density <- outer(tau_grid, lambda_grid, gstar)
op <- par(bg = "white") 
z_max <- max(density, na.rm=TRUE)
persp(
  tau_grid, 
  lambda_grid, 
  density, 
  zlim=c(0, 1.1*z_max),
  xlab="tau",
  ylab="lambda",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

This is the joint prior distribution of $\tau$ and $\lambda$. 

It is hard to see at this scale, but the prior distribution is higher for small values of $\lambda$ when $\tau$ is very large or very small. This becomes more obvious if you remove the middle of the surface and rescale the Z-axis.

```{r normal-hedging-prior-27, eval=FALSE}
density <- outer(tau_grid, lambda_grid, gstar)
density[16:44, ] <- NA
op <- par(bg = "white")
z_max <- max(density, na.rm=TRUE)
persp(
  mu_grid, 
  lambda_grid, 
  density, 
  zlim=c(0, 1.1*z_max),
  xlab="tau",
  ylab="lambda",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

The likelihood function is constant across all values of $\tau$, so its surface looks like the following.

```{r normal-hedging-prior-28, eval=FALSE}
lambda_grid <- seq(0, 1, length=60)
tau_grid <- seq(0, x_max, length=60)


fstar <- function(mu, lambda) {f(xbar, mu, tau0)^n}
likelihood <- outer(mu_grid, lambda_grid, fstar)
z_max <- max(likelihood, na.rm=TRUE)
persp(
  mu_grid, 
  lambda_grid, 
  likelihood, 
  zlim=c(0, 1.1*z_max),
  xlab="mu",
  ylab="lambda",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

The product of the prior and the likelihood is proportional to the posterior distribution.

```{r normal-hedging-prior-29, eval=FALSE}
lambda_grid <- seq(0, 1, length=60)
mu_grid <- seq(0, x_max, length=60)


hstar <- function(mu, lambda) {g(mu, mu0, lambda*(tau0-tauF) + tauF)*f(xbar, mu, tauF)^n}
density <- outer(mu_grid, lambda_grid, hstar)
z_max <- max(density, na.rm=TRUE)
persp(
  mu_grid, 
  lambda_grid, 
  density, 
  zlim=c(0, 1.1*z_max),
  xlab="mu",
  ylab="lambda",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

A contour plot shows where the probability is concentrated. The labels on the contour lines represent how much probabilty is found inside the contour line.

```{r normal-hedging-prior-30, eval=FALSE}
density <- density / sum(density, na.rm=TRUE)
density %>%
  as.vector %>%
  sort %>%
  rev -> sorted_density
cumulative_p <- 0
for (p in sorted_density) {
  cumulative_p <- p + cumulative_p
  density[density==p] <- cumulative_p
}
contour(
  mu_grid, 
  lambda_grid, 
  density,
  labcex=1,
  levels=c(50, 75, 90)/100,
  xlab="tau",
  ylab="lambda")
```

You can see that the hedging hyperprior did its job. The most probable values of $\lambda$ are small and the prior mean (`r tau0`) is largely ignored in favor of the mean of the data (`r "??"`).

[jor1]: https://people.eecs.berkeley.edu/~jordan/courses/260-spring10/lectures/lecture5.pdf
[mur1]: https://www.cs.ubc.ca/~murphyk/Papers/bayesGauss.pdf
[wik1]: https://en.wikipedia.org/wiki/Completing_the_square

