---
title: "Check cross-links between new and old site"
author: "Steve Simon"
date: "Created 2024-06-25"
execute:
  echo: false
format:
  html:
    embed-resources: true
---
  
This program compares links between the new and old websites.

```{r setup}
size_limit <- 999
verbose <- (size_limit > 0)
suppressMessages(suppressWarnings(library(glue)))
suppressMessages(suppressWarnings(library(magrittr)))
suppressMessages(suppressWarnings(library(readr)))
suppressMessages(suppressWarnings(library(stringr)))
suppressMessages(suppressWarnings(library(tidyverse)))

load(file="../data/check_new.RData")
n_new <- nrow(new)
load(file="../data/check_old.RData")
n_old <- nrow(old)

inc <- function(x) {
  n <- length(x)
  m <- c(FALSE, x[-1]==x[-n])
  x[m] <- " "
  x
}
```

```{r test-inc, echo=FALSE, eval=FALSE}
test1 <- rep(letters[1:4], 1:4)
test2 <- inc(test1)
data.frame(test1, test2)
```

## pages with no links from new site to old or blog site

```{r no-link-www-1}
new %>% 
  filter(link_miss==1) %>%
  select(dir, page) -> pip1 
n <- nrow(pip1)
pct <- round(100*n/n_new)
msg <- glue("{n} / {n_new} = {pct}%")
```

### `r msg`

```{r no-link-www-2}
if (verbose) {
  pip1 %>%
    mutate(dir=inc(dir)) %>%
    head(size_limit)
}
```

## pages with bad links from new site to old site

```{r bad-link-www-1}
old %>%
  rename(back_link=link_page) %>%
  rename(link_page=page) %>%
  rename(link_dir=dir) %>%
  select(link_dir, link_page, back_link, stub) -> mrg
new %>% 
  filter(link_miss==0) %>%
  filter(link_site=="www") %>%
  left_join(mrg, by=c("link_dir", "link_page")) %>%
  select(
    dir, 
    page, 
    link_dir, 
    link_page, 
    back_link,
    stub) -> pip2
pip2 %>%
  filter(is.na(stub)) %>%
    select(-stub) -> pip3 
n <- nrow(pip3)
pct <- round(100*n/n_new)
msg <- glue("{n} / {n_new} = {pct}%")
```

### `r msg`

```{r bad-link-www-2}
if (verbose) {
  pip3 %>%
    mutate(dir=inc(dir)) %>%
    head(size_limit)
}
```


## pages with good links from new site to old site

```{r good-link-www-1}
pip2 %>%
  filter(!is.na(stub))  -> pip4 
n <- nrow(pip4)
pct <- round(100*n/n_new)
msg <- glue("{n} / {n_new} = {pct}%")
```

### `r msg`

```{r good-link-www-2}
if (verbose) {
  pip4 %>%
    mutate(dir=inc(dir)) %>%
    head(size_limit)
}
```

## pages with links from new site to blog site

Note: there is no easy way for a program to check whether the link to my blog site is bad or good.

```{r link-blog-1}
new %>%
  filter(link_site=="blog") %>%
  filter(link_miss==0) -> pip5 
n <- nrow(pip5)
pct <- round(100*n/n_new)
msg <- glue("{n} / {n_new} = {pct}%")
```

### `r msg`

```{r link-blog-2}
if (verbose) {
  pip5 %>%
    mutate(dir=inc(dir)) %>%
    head(size_limit)
}
```

## pages with no links from old site to new site

```{r no-link-old-1}
old %>% 
  filter(link_miss==1) %>%
  select(dir, page) -> pip6 
n <- nrow(pip6)
pct <- round(100*n/n_old)
msg <- glue("{n} / {n_old} = {pct}%")
```

### `r msg`

```{r no-link-old-2}
pip6 %>%
  mutate(dir=inc(dir)) %>%
  head(size_limit)
```

## pages with bad links from old site to new site

```{r bad-link-old-1}
new %>%
  rename(back_link=link_page) %>%
  rename(link_page=page) %>%
  select(link_page, back_link, update) -> mrg

old %>%
  filter(link_miss==0) %>%
  left_join(mrg, by="link_page") %>%
  select(dir, page, link_page, update) -> pip7

pip7 %>%
  filter(is.na(update)) %>%
  select(-update)  -> pip8

n <- nrow(pip8)
pct <- round(100*n/n_old)
msg <- glue("{n} / {n_old} = {pct}%")
```

### `r msg`

```{r bad-link-old-2}
if (verbose) {
  pip8 %>%
    mutate(dir=inc(dir)) %>%
    head(size_limit)
}
```


## pages with good links from old site to new site

```{r good-link-old-1}
pip7 %>%
  filter(!is.na(update)) -> pip9 
n <- nrow(pip9)
pct <- round(100*n/n_old)
msg <- glue("{n} / {n_old} = {pct}%")
```

### `r msg`

```{r good-link-old-2}
if (verbose) {
  pip9 %>%
    mutate(dir=inc(dir)) %>%
    head(size_limit)
}
```

