---
title: "Nonlinear regression for the difference of two exponentials"
author: "Steve Simon"
source: "http://blog.pmean.com/difference-of-two-exponentials/"
date: "2015-10-05"
category: 
- Blog post
tags:
- incomplete page
- nonlinear regression
- R software
output: html_document
---

I wanted to provide an overview of how you analyze a classic nonlinear regression model. It is a difference of two exponential functions. This nonlinear function is used commonly in pharmocokinetic models and is a simply way to model the oral administration of a drug. I want to show how the model works in a mathematical sense and then how you fit it using R.

<!---More--->

Here are some simple examples of nonlinear regression. We will use the built-in data set Theoph for all our examples.

```{r}
head(Theoph)
```

```{r}
tail(Theoph)
```

### look at just the first patient.

You can find an [earlier version][sim1] of this page on my [blog][sim2].

[sim1]: http://blog.pmean.com/difference-of-two-exponentials/
[sim2]: http://blog.pmean.com
