---
title: "Examining relationships in R"
author: "Steve Simon"
source: "http://blog.pmean.com/relationships-in-r/"
date: "2015-04-03"
categories:
- Blog post
tags:
- R software
- Rmarkdown graphs
output: html_document
page_update: complete
---

I'm giving a talk for the Kansas City R Users Group on how to get a
preliminary impression of relationships between pairs of variables. Here
is the R code and output that I will use.

<!---More--->

### Simple measures of association

There are several different ways of measuring bivariate relationships in
a descriptive fashion prior to data analysis. The methods can be largely
grouped into measures of relationship between two continuous variables,
two categorical variables and measures of a relationship between a
categorical variable and a continuous variable.

```{r relationships-in-r-01}
suppressMessages(suppressWarnings(library(tidyverse)))
fn <- "https://raw.githubusercontent.com/pmean/introduction-to-SAS/master/data/housing.txt"
home <- read_tsv(fn, na=".")
home
```

The best graphical summary of two continuous variables is a scatterplot.
You should include a smoothing curve or spline model to the graph to
emphasize the general trend and any departures from linearity.

```{r relationships-in-r-02}
plot(home$sqft,home$price)
lines(lowess(home$price~home$sqft))
```



```{r relationships-in-r-03}
sb <- is.finite(home$age)
plot(home$age[sb],home$price[sb])
lines(lowess(home$price[sb]~home$age[sb]))
```



The best numeric summary of two continuous variables is a correlation
coefficient.

```{r relationships-in-r-04}
cor(home[,c("price","sqft","age")],use="pairwise.complete.obs")
```

Correlations should always be rounded to two or maybe even just one
significant digit.

```{r relationships-in-r-05}
round(cor(home[,c("price","sqft","age")],use="pairwise.complete.obs"),1)
```

Anything larger than 0.7 or smaller than -0.7 is a strong linear
relationship. Anything between 0.3 and 0.7 or between -0.3 and -0.7 is a
weak linear relationship. Anything between -0.3 and 0.3 represents
little or no linear relationship.

The best graphical summary between a continuous variable and a
categorical variable is a boxplot.

```{r relationships-in-r-06}
boxplot(home$price~home$features)
```


```{r relationships-in-r-07}
boxplot(home$price~home$northeast)
```



```{r relationships-in-r-08}
boxplot(home$price~home$custom_build)
```


```{r relationships-in-r-09}
boxplot(home$price~home$corner_lot)
```




If your categorical variable is binary, you can also use a scatterplot.
The binary variable goes on the y axis and a trend line is critical.

```{r relationships-in-r-10}
plot(home$price,home$northeast)
lines(lowess(home$northeast~home$price))
```




```{r relationships-in-r-11}
plot(home$price,home$custom_build)
lines(lowess(home$custom_build~home$price))
```




```{r relationships-in-r-12}
plot(home$price,home$corner_lot)
lines(lowess(home$corner_lot~home$price))
```



You can also compute a correlaton between a binary variable and a
categorical variable. It is equivalent to the point-bisearial
correlation.

```{r relationships-in-r-13}
round(cor(home[,c("northeast","custom_build","corner_lot")],home$price),1)
```

Let's save the display of a relationship involving two categorical
variables until another day.


